Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-06 Paper/intro.tex

The application problem considered in this work is the
H$_2$/O$_2$ reaction mechanism~\cite{Yetter:1991}. This mechanism has gained
a lot of attention as a potential source of clean energy
locomotive applications~\cite{Das:1996}, and more recently in fuel cells~\cite{Loges:2008}. 
The mechanism involves 19
reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in
table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
The equivalence ratio ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the fuel (H$_2$)-oxidizer
(O$_2$) ratio at given condition to the same quantity under stoichiometric
conditions. In this study, computations were performed at fuel-rich conditions,
$\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of
temperature increase to exceed a given threshold, regarded as \emph{ignition delay}
is recorded.

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on ignition delay. The $A_i$'s are considered to be
uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem high-dimensional and computationally challenging. 
For the purpose of dimension reduction, the active subspace is evaluated using
two computational strategies: a gradient-based approach and  
a gradient-free approach. These strategies are based on algorithms 1.1 and 1.2 
in~\cite{Constantine:2015}. As mentioned earlier, the gradient information of the model
output is needed in both cases. However, in the case of gradient-free approach,
the gradient is approximated using a linear regression-fit to the model output.
Hence, the amount of computational effort is reduced in this case. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated at each iteration until convergence is accomplished as discussed
later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global sensitivity
measures associated with uncertain inputs of a model. Uncertainty along the inputs with
a relatively small estimate of the sensitivity measure can be ignored for the purpose of
reducing the computational effort associated with uncertainty quantification (UQ). The
Sobol' indices are commonly used for this purpose~\cite{Sobol:2001}. However, obtaining
converged estimates of the Sobol' indices typically requires tens of thousands of model
runs and is therefore impractical in situations involving intensive model simulations.
Recently, the derivative-based global sensitivity measures~(DGSMs) have been 
developed to approximate the upper bound on the Sobol' indies with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. In Section~\ref{sec:assa}, we show
that the DGSMs can in fact be approximated using the components of the eigenvectors
in the active subspace.

Main contributions of this paper are as follows: (1) Mathematical linkages between the
global sensitivity measures: Total Sobol index, DGSMs, and the activity scores are
established; (2) An efficient gradient-free iterative approach based on local linear
approximation of the model output is shown to yield consistent results with the
gradient-based approach. The two approaches were implemented to compute
the active subspace for a 19-dimensional H$_2$/O$_2$ kinetics problem with 
uncertain pre-exponents in the rate law for individual reactions. (3) The gradient-free
approach is finally implemented to discover an active subspace for the 33-dimensional
H$_2$/O$_2$ reaction kinetics application in which case the pre-exponents as well
as the activation energies are considered to be uncertain.

 The remaining of this paper is organized as follows. In Section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In Section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In Section~\ref{sec:method}, a systematic framework
 for computing the active subspace using the gradient-based, and the gradient-free approaches is presented. 
The two approaches are implemented to the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. Our findings illustrate the consistency in the results based on
the two approaches thereby encouraging the use of the gradient-free approach. In Section~\ref{sec:app},
the gradient-free approach is implemented to the 33-dimensional
H$_2$/O$_2$ reaction kinetics application by accounting for the uncertainty in $E_{a,i}$'s as well. The
resulting active subspace is found to be 1-dimensional indicating enormous potential for reduction in
computational effort pertaining to UQ, in this case. 
Finally, we draw important conclusions based on the various aspects of this work in Section~\ref{sec:conc}.

>>>>

The application problem considered in this work is the
H$_2$/O$_2$ reaction mechanism~\cite{Yetter:1991}. This mechanism has gained
a lot of attention as a potential source of clean energy for
locomotive applications~\cite{Das:1996}, and more recently in fuel cells~\cite{Loges:2008}. 
The mechanism involves 19
reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in
table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
The equivalence ratio ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the ratio of the fuel (H$_2$)
and oxidizer (O$_2$) at a given condition to the same quantity under stoichiometric
conditions. In this study, computations were performed at fuel-rich conditions,
$\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of
temperature increase to exceed a given threshold, regarded as \emph{ignition delay}
is recorded.

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s are considered to be
uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem high-dimensional and computationally challenging. 
To address this challenge, we focus our efforts on reducing the dimensionality
of the problem by computing the active subspace using
two computational strategies: a gradient-based approach and  
a gradient-free approach. These strategies are based on algorithms 1.1 and 1.2 
in~\cite{Constantine:2015}. As mentioned earlier, the gradient information of the model
output is needed in both cases. However, in the case of gradient-free approach,
the gradient is approximated using a linear regression-fit to the model output.
Hence, the amount of computational effort is reduced in this case. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated, and gradient evaluations are performed at each iteration until 
a reasonably accurate approximation to the active subspace is achieved, as
discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global sensitivity
measures associated with uncertain inputs of a model. Depending upon the estimates of
the sensitivity measures, only the important inputs are varied for the purpose of
uncertainty quantification (UQ). The
Sobol' index is a commonly used global sensitivity measure~\cite{Sobol:2001}. However, obtaining
converged estimates of the Sobol' indices typically requires tens of thousands of model
runs and is therefore impractical in situations involving intensive model simulations.
Recently, the derivative-based global sensitivity measures~(DGSMs) were 
developed to approximate the upper bound on the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. In Section~\ref{sec:assa}, we show
that the DGSMs can in fact be approximated using the components of the eigenvectors
in the active subspace by estimating the so-called activity 
scores~\cite{Diaz:2016,Constantine:2017}.

Main contributions of this paper are as follows: (1) Mathematical linkages between the
global sensitivity measures: Total Sobol' index, DGSMs, and the activity scores are
established; (2) An efficient gradient-free iterative approach using regression-based local linear
approximation of the model output is shown to yield consistent results with the
gradient-based approach. The two approaches were implemented to compute
the active subspace for a 19-dimensional H$_2$/O$_2$ kinetics problem with 
uncertain pre-exponents in the rate law for individual reactions. (3) The gradient-free
approach is finally implemented to discover an active subspace for the 33-dimensional
H$_2$/O$_2$ reaction kinetics application in which case the pre-exponents as well
as the activation energies are considered to be uncertain.

 The remaining of this paper is organized as follows. In section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In section~\ref{sec:method}, a systematic framework
 for computing the active subspace using the gradient-based, and the gradient-free approaches is presented. 
The two approaches are implemented to the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. Our findings illustrate the consistency in the results based on
the two approaches thereby encouraging the use of the gradient-free approach in high-dimensional settings.
 In section~\ref{sec:app},
the gradient-free approach is implemented to the 33-dimensional
H$_2$/O$_2$ reaction kinetics application by accounting for the uncertainty in $E_{a,i}$'s as well. The
resulting active subspace is found to be 1-dimensional indicating enormous potential for reduction in
computational effort pertaining to UQ, in this case. 
Finally, we summarize the results and include a discussion on this work in section~\ref{sec:conc}.
